### 2016-05-19 Paper_v2.tex
Updates, added text

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamic modeling of human complement}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Wei Dai$^{\#}$, Mason Minot$^{\#}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Dynamic Modeling of Human Complement System using Reduced Ordered Models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{???????}}
\vspace{0.5in}
{\par \centering $^{\#}$ Denotes equal contribution}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Fill me in.

@@

\section*{Introduction}
Complement is a central part of innate immunity and plays a very significant role in regulating the inflammatory response. Complement was first discovered in the 1890s where it was found to 'complement' the bactericidal activity of natural antibodies. Complement is mediated through a set of approximately 30-35 soluble and cell surface proteases. The central process in complement activation involves the formation of Membrane Attack Complex (MAC) and a protein called C5a. C5a acts as a bridge between innate and adaptive immunity and plays a very important role in regulating inflammation and coagulation.  Complement activation takes places through three different pathways: the alternate, the classical and the lectin. Each of these pathways involves a different initiator signal that leads to the formation of a serine protease called C5 convertase which cleaves an inactive protein called C5 to form C5a and C5b. The classical pathway is triggered when antibodies form complexes with foreign antigens or other pathogens. A multimeric protein complex C1 binds to the antigen-antibody complex and undergoes a conformational change. This activated complex cleaves proteins C4 and C2 to C4a, C4b, C2a and C2b respectively. C4a and C2b combine to form a protease C4bC2a also known as the classical C3 convertase. The lectin pathway is initiated through the binding of L-ficolin or Mannose Binding Lectin (MBL) to the carbohydrates on the surfaces of bacterial pathogens. This bound complex in turn cleaves C4 and C2 and leads to the production of C4bC2a. The alternate pathway involves a 'tickover' mechanism in which a protein called C3 is hydrolyzed to form C3b. In presence of foreign pathogens C3b binds to these surfaces and recruits  additional factors called factor B and factor D that lead to the formation of alternate C3 convertase - C3bBb. The formation of classical and alternate C3 convertases on bacterial surfaces is followed by the formation of proteases called C5 convertases. The classical and alternate C3 convertases recruit C3, Factor B and Factor D to form classical C5 convertase (C4bC2aC3b) and alternate C5 convertase (C3bBbc3B) respectively. The C5 convertases then cleave C5 to form C5a and C5b respectively. 	The cleavage of C5 is followed by a series of sequential cleavages of proteins C6, C7, C8 and C9 that combine with C5b to form the MAC complex.  The activation of complement and formation of C5a and MAC complex is regulated at different points through a number of plasma and host cell proteins.  C4 binding protein (C4BP), ---- (NEEDS TO BE FILLED)

Given the complexity and importance of complement, developing mathematical models of complement are crucial to understanding its dynamics. Complement models have typically been formulated as linear or non-linear Ordinary Differential Equation (ODE) systems. Hirayama  et al. (ref) used a system of linear ODEs to model the classical pathway of complement. Korotaevskiy and co-workers (ref) built a theoretical model of complement using a system of non-linear ODEs that included classical, lectin and alternate pathways. However both these studies involve no validation studies with experimental data. Liu et al used analyzed the formation of classical and lectin C3 convertases and the regulatory role of C4BP using a system of 45 non-linear ODEs with 85 parameters.  Recently, Zewde and co-workers built a detailed mechanistic model of alternative complement activation was built using 107 ODEs and 74 kinetic parameters (Ref). This model delineated the response of complement on a host cell and a foreign antigen. However, these previous models were largely based upon mechanistic knowledge.  However given the complexity of complement and its interactions with other networks like coagulation, autonomous nervous system, adaptive immunity it is unfeasible and computationally expensive to build such large mechanistic models. In addition is much more difficult to experimentally interrogate the response of various complement proteins under different conditions. This also presents with the problem of estimation of a large number of parameters with little or no experimental data. Thus there exists a need to reduce the mechanistic complexity while capturing dynamics of complement accurately.

In this study, we formulated a model of the human complement system. [FILL ME IN].

@@

\subsection*{Formulation of a reduced order complement model.}
We developed a reduced order extrinsic human complement network consisting of the most crucial steps of the human complement system (Fig. \ref{fig-schematic}). The core of our model was based upon the experimental measurements of Morad and coworker's earlier work \cite{morad2015time}, we only consider the activation of complement system through the alternate and the lectin pathways. In doing so we aim to capture a complex biological phenomenon using a few simple ordinary differential equations.  A trigger event initiates the lectin pathway in the presence of zymosan, which activates the cleavage of $C2$ and $C4$ into $C2a$ and $C2b$, and $C4a$ and $C4b$ respectively. Classical Pathway ($CP$) $C3 convertase$ ($C4aC2b$) is by the combination of $C4a$ and $C2b$, which catalyzes the cleavage of $C3$ into $C3a$ and $C3b$. Similarly, the activation of the alternative pathways happens through the spontaneous hydrolysis of $C3$ which facilitates the cleavage of $C3$. $C3b$ then could combine with with $C3$ to form alternate pathway ($AP$) $C3 convertase$. The both  versions of the $C3 convertase$ catalyzes the cleavage of $C3$ into $C3a$ and $C3b$, and $C3b$ can then combine with either $CP$ or $AP C3 convertase$ to form $C5 convertase$, $CP or AP$ respectively that is responsible for the cleavage of $C5$ to $C5a$ and $C5b$. Lectin pathway activation was approximated using a combination of saturation kinetics and Hill-like function control functions.These control coefficients then modified the rates of model processes at each time step. Hill-like transfer functions $0 \le \mathbf{f (Z)}  \le 1$ quantified the contribution of components upon a target process, in this study, $\mathbf{Z}$ represents the abundance of the initiator. Taken together, while the reduced order human complement model encodes significant biological complexity, it is highly compact (consisting of only 18 differential equations). Thus, it will serve as an excellent proof
of principle example to study the reduction of a highly complex human subsystem.

\subsection*{Model were estimated using dynamically dimensioned search.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. We estimated
kinetic and control parameters in a hierarchical fashion using  two $in~vitro$ time-series human complement data sets with and without zymosan present. The residual between simulation and experimental measurements were minimized  using dynamically dimensioned search (DDS). An initial parameter set was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized the residual. Knowing that the kinetic and control parameters of the lectin pathway does not affect the dynamics of the alternate pathway, we used a hierarchal approach that estimated the parameters for the alternative pathway and lectin pathway separately. For the alternative pathway, we utilized the time-course experimental measurements of Morad and coworkers \cite{morad2015time} of $C3a$ and $C5a$ in the absence of zymosan and only allowed the alternative parameters to vary (Fig. \ref{fig-fit} A and B). The estimated alternate parameters was then fixed for the determination of lectin pathway parameters. The training for the lectin parameters, we used the experimental measurements of $C3a$ and $C5a$ in the presence of $1$ g of zymosan published by Morad et al \cite{morad2015time} (Fig. \ref{fig-fit} C and D).

The reduced human complement model captured the behavior of the alternative and lectin pathways through the time-course abundance of $C3a$ and $C5a$ (Fig. \ref{fig-fit}). However we were not able to capture the curvature of the $C5a$ alternate (Fig. \ref{fig-fit}), the decreasing slope of the experimental measurements may be an indication of the decreasing cofactors that are required for the spontaneous hydrolysis in the alternative pathway, which we neglected.  Taken together, the model identification results suggested that our reduced order approach could reproduce a panel of lectin pathway intiation data sets in the neighborhood of physiological factor and inhibitor concentrations. However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.

\subsection*{Validation of the reduced order human complement model.}
We tested the predictive power of the reduced order human complement model with validation data sets not used during model training. Six validation data sets were used, three for $C3a$ and $C5a$ respectively at different zymosan concentrations. All kinetic and control parameters were
fixed for the validation simulations. The reduced order model predicted the $C3a$ and $C5a$ time-course profiles at a qualitative level (Fig. \ref{fig-prediction}).

@@

@@

\section*{Supplemental materials.}
\subsubsection*{Model equations.}
Applying the general framework to the reduced coagulation network resulted in 18 ordinary differential equations, 12 rate equations, and two control equations:
\begin{eqnarray}
	\frac{dx_{1}}{dt} & =& - r_{init,c4}f_{init,c4} \\ 								%C4
	\frac{dx_{2}}{dt} &=& - r_{init,c2}f_{init,c2} \\  								%C2
	\frac{dx_{3}}{dt} &=&  r_{init,c4}f_{init,c4} \\ 								%C4a
	\frac{dx_{4}}{dt} &=& r_{init,c4}f_{init,c4} - r_{cp,c3c,form} \\ 					%C4b
	\frac{dx_{5}}{dt} & = & r_{init,c2}f_{init,c2} - r_{cp,c3c,form} \\ 					%C2a
	\frac{dx_{6}}{dt} &=& r_{init,c2}f_{init,c2} \\ 								%C2b
	\frac{dx_{7}}{dt} &=& r_{c3,basal} - r_{cp,c3c,c3b} - r_{ap,c3c,c3b}\\ 				%C3
	\frac{dx_{8}}{dt} &=& r_{c3,basal}  + r_{cp,c3c,c3b} + r_{ap,c3c,c3b}  - k_{deg,c3a}*C3a \\		%C3a
	\frac{dx_{9}}{dt} &=& r_{c3,basal}  + r_{cp,c3c,c3b} + r_{ap,c3c,c3b}  -  r_{ap,c3c,form} \\		%C3b
	\frac{dx_{10}}{dt} &=& r_{cp,c3c,form} - r_{cp,c5c,form} -  r_{cp,c3c,c4bp,inhib} \\			%CP C3C
	\frac{dx_{11}}{dt} &=& r_{ap,c3c,form} - r_{ap,c5c,form} -  r_{ap,c3c,factorH,inhib} \\			%AP C3C
	\frac{dx_{12}}{dt} &=& r_{cp,c5c,form} - r_{cp,c5c,c4bp,inhib} \\							%CP C5C
	\frac{dx_{13}}{dt} &=& r_{cp,c5c,form}  \\											%AP C5C
	\frac{dx_{14}}{dt} &=& - r_{cp,c5c,c5b} - r_{ap,c5c,c5b} \\								%C5
	\frac{dx_{15}}{dt} &=&  r_{cp,c5c,c5b} + r_{ap,c5c,c5b} - k_{deg,c5a} \\					%C5a
	\frac{dx_{16}}{dt} &=& r_{cp,c5c,c5b} + r_{ap,c5c,c5b} \\								%C5b
	\frac{dx_{17}}{dt} &=& - r_{cp,c3c,c4bp,inhib} - r_{cp,c5c,c4bp,inhib} \\  				         %C4BP
	\frac{dx_{18}}{dt} &=& - r_{ap,c3c,factorH,inhib} \\									%Factor H
\end{eqnarray}
where \begin{multline}
f{x} = f(Zymosan,C2,C2a,C2b,C4,C4a,C4b,C3,C3a,C3b,CP C3C Convertase,... \\
           AP C3 Convertase,CP C5 Convertase,AP C5 Convertase,C5,C5a,... \\
           ... C5b,C4BP,Factor H)^{T}
\end{multline}

The various rate and control equations are given by:
\begin{eqnarray}
	r_{init,c4} &=& \frac{k_{init,c4}*C4}{K_{init,c4,s} + C4} \\
	r_{init,c2} &=& \frac{k_{init,c2}*C2}{K_{init,c2,s} + C2} \\
	f_{init,c4} &=& \frac{Zymo^{\eta_{zymo,c4}}}{Zymo^{\eta_{zymo,c4}} + alpha_{c2}^{\eta_{zymo,c4}}} \\
	f_{init,c2} &=& \frac{Zymo^{\eta_{zymo,c2}}}{Zymo^{\eta_{zymo,c2}} + alpha_{c4}^{\eta_{zymo,c2}}} \\
	r_{c3,basal} &=& k_{c3b,basal}*C3 \\
	r_{cp,c3c,c3b} &=& \frac{k_{cp,c3c}*CP,C3C*C3^{\eta_{cp,c3c}}}{K_{cp,c3c,s}^{\eta_{cp,c3c}} + C3^{\eta_{cp,c3c}}}  \\
	r_{ap,c3c,c3b} &=& \frac{k_{ap,c3c}*AP,C3C*C3}{K_{ap,c3c,s} + C3} \\
	r_{cp,c3c,form} &=& k_{cp,c3c,form}*C4b*C2a \\
	r_{ap,c3c,form} &=& k_{ap,c3c,form}*C4b*C2a \\
	r_{cp,c3c,c4bp,inhib} &=& k_{cp,c3c,inhib}*CP,C3C*C4b*C4BP \\
	r_{ap,c3c,factorH,inhib} &=& k_{ap,c3c,inhib}*AP,C3C*FactorH \\
	r_{cp,c5c,form} &=& k_{cp,c5c,form}*CP,C3C*C3b \\
	r_{ap,c5c,form} &=& k_{ap,c5c,form}*AP,C3C*C3b \\
	 r_{cp,c5c,c5b} &=& \frac{k_{cp,c5c}*CP,C5C*C5^{\eta_{cp,c5c}}}{K_{cp,c5c,s}^{\eta_{cp,c5c}} + C5^{\eta_{cp,c5c}}} \\
	 r_{ap,c5c,c5b} &=& \frac{k_{ap,c5c}*AP,C5C*C5}{K_{ap,c5c,s} + C5} \\
	 r_{cp,c5c,c4bp,inhib} &=& k_{cp,c5,inhib}*CP,C5C*C4BP
\end{eqnarray}

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Reduced order modeling and analysis of the human complement subsystem}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Wei Dai$^{\#}$, Mason Minot$^{\#}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order model of complement}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS~ONE}}
\vspace{0.5in}
{\par \centering $^{\#}$ Denotes equal contribution}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Complement is a central part of innate immunity and plays a very significant role in regulating the inflammatory response.
In this study, we present an effective biochemical network modeling approach in building a reduced order model to study a complex biochemical network, the human complement system. The key innovation of our approach is the use of simple equations to capture the behavior of a complex biochemical network.  We developed a hybrid modeling approach which combines ODEs and logical rules to model biochemical processes for which a complete mechanistic understanding is missing. Using this modeling framework we  incorporated computational simulation and experimental validation of the lectin and alternative pathway to examine C3 and C5 convertase formation and activity. The reduced order model consisted of only 18 differential equations with 28 kinetic and control parameters. Thus, the model was several orders of magnitude smaller and includes more pathways than comparable purely ODE models in the literature. We estimated the model parameters from in vitro human complement time-course experiments of C3a and C5a from Morad and coworkers \cite{morad2015time}, in the presence and absence of zymosan, and without the classical pathway. We then compared the model predictions with C3a and C5a data sets, for alternative and lectin pathway dynamics, that were not used for model training. Once validated, we performed sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. The reduced order hybrid approach produced a surprisingly predictive human complement model, much similar to the study on human coagulation using the same modeling framework \cite{sagar2015dynamic}. Taken together, the combined analysis of alternate and lectin pathways along with the incorporation of the downstream reactions involving C5 convertase elucidated new insight into roles of governing parameters that mediate the complement system. Deeply understanding how the main governing parameters change with respective to the different combinations of pathway activation will greatly aid in development of drugs for strategic therapeutic targets. Due to the low computational cost relative to the existing models and accuracy of our predictions, we believe that our reduced order complement network is the first step towards building a computation toolbox for treatment analysis or clinical screening in the medical field.

@@

\section*{Introduction}
Complement is a central part of innate immunity and plays a very significant role in regulating the inflammatory response. Complement was first discovered in the 1890s where it was found to 'complement' the bactericidal activity of natural antibodies. Complement is mediated through a set of approximately 30-35 soluble and cell surface proteases. The central process in complement activation involves the formation of Membrane Attack Complex (MAC) and a protein called C5a. C5a acts as a bridge between innate and adaptive immunity and plays a very important role in regulating inflammation and coagulation.  Complement activation takes places through three different pathways: the alternate, the classical and the lectin. Each of these pathways involves a different initiator signal that leads to the formation of a serine protease called C5 convertase which cleaves an inactive protein called C5 to form C5a and C5b. The classical pathway is triggered when antibodies form complexes with foreign antigens or other pathogens. A multimeric protein complex C1 binds to the antigen-antibody complex and undergoes a conformational change. This activated complex cleaves proteins C4 and C2 to C4a, C4b, C2a and C2b respectively. C4a and C2b combine to form a protease C4bC2a also known as the classical C3 convertase. The lectin pathway is initiated through the binding of L-ficolin or Mannose Binding Lectin (MBL) to the carbohydrates on the surfaces of bacterial pathogens. This bound complex in turn cleaves C4 and C2 and leads to the production of C4bC2a. The alternate pathway involves a 'tickover' mechanism in which a protein called C3 is hydrolyzed to form C3b. In presence of foreign pathogens C3b binds to these surfaces and recruits  additional factors called factor B and factor D that lead to the formation of alternate C3 convertase - C3bBb. The formation of classical and alternate C3 convertases on bacterial surfaces is followed by the formation of proteases called C5 convertases. The classical and alternate C3 convertases recruit C3, Factor B and Factor D to form classical C5 convertase (C4bC2aC3b) and alternate C5 convertase (C3bBbc3B) respectively. The C5 convertases then cleave C5 to form C5a and C5b respectively. 	The cleavage of C5 is followed by a series of sequential cleavages of proteins C6, C7, C8 and C9 that combine with C5b to form the MAC complex.

The activation of complement and formation of C5a and MAC complex is regulated at different points through a number of plasma and host cell proteins.
The initiation of the classical pathway through the attachment of C1 to an antibody is controlled by the C1 Inhibitor (C1-Inh), a protease inhibitor belonging to the serpin superfamily. C1-Inh irreversibly binds to and deactivates the active subunits of component C1 to prevent spontaneous fluid phase and chronic activation of complement \cite{walker1995complement}. The serum and host-tissue regulation of the upstream elements of the complement system is also achieved through the binding of C4 binding protein (C4BP) to C4b and through the binding of factor H to C3b \cite{blom2001structural}. These proteins are also capable of binding their respective components in the convertase form. Membrane cofactor protein (MCP or CD46) possesses a cofactor activity for C4b and C3b, which protects the host from self-activation of complement \cite{riley2004cd46}.  Delay accelerating factor (DAF or CD55) is able to recognize and dissociate both convertases \cite{lukacik2004complement}. MAC is inhibited by vitronectin and clusterin in the plasma and CD59 at the host surface.
Proteins C3a, C4a, and C5a are inactivated or reduced in activity by carboxypeptidase-N \cite{liszewski1995control}.

Given the complexity and importance, developing mathematical models of complement are crucial to understanding its dynamics. Complement models have typically been formulated as linear or non-linear Ordinary Differential Equation (ODE) systems. Hirayama  et al. (ref) used a system of linear ODEs to model the classical pathway of complement. Korotaevskiy and co-workers (ref) built a theoretical model of complement using a system of non-linear ODEs that included classical, lectin and alternate pathways. However both these studies involve no validation studies with experimental data. Liu et al used analyzed the formation of classical and lectin C3 convertases and the regulatory role of C4BP using a system of 45 non-linear ODEs with 85 parameters.  Recently, Zewde and co-workers built a detailed mechanistic model of alternative complement activation was built using 107 ODEs and 74 kinetic parameters (Ref). This model delineated the response of complement on a host cell and a foreign antigen. However, these previous models were largely based upon mechanistic knowledge.  However given the complexity of complement and its interactions with other networks like coagulation, autonomous nervous system, adaptive immunity it is unfeasible and computationally expensive to build such large mechanistic models. In addition is much more difficult to experimentally interrogate the response of various complement proteins under different conditions. This also presents with the problem of estimation of a large number of parameters with little or no experimental data. Thus there exists a need to reduce the mechanistic complexity while capturing dynamics of complement accurately.

In this study, we present an effective biochemical network modeling approach in building a reduced order model to study a complex biochemical network, the human complement system. The key innovation of our approach is the use of simple equations to capture the behavior of a complex biochemical network.  We developed a hybrid modeling approach which combines ODEs and logical rules to model biochemical processes for which a complete mechanistic understanding is missing. Using this modeling framework we  incorporated computational simulation and experimental validation of the lectin and alternative pathway to examine C3 and C5 convertase formation and activity. The reduced order model consisted of only 18 differential equations with 28 kinetic and control parameters. Thus, the model was several orders of magnitude smaller and includes more pathways than comparable purely ODE models in the literature. We estimated the model parameters from in vitro human complement time-course experiments of C3a and C5a from Morad and coworkers \cite{morad2015time}, in the presence and absence of zymosan, and without the classical pathway. We then compared the model predictions with C3a and C5a data sets, for alternative and lectin pathway dynamics, that were not used for model training. Once validated, we performed sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. The reduced order hybrid approach produced a surprisingly predictive human complement model, much similar to the study on human coagulation using the same modeling framework \cite{sagar2015dynamic}. Taken together, the combined analysis of alternate and lectin pathways along with the incorporation of the downstream reactions involving C5 convertase elucidated new insight into roles of governing parameters that mediate the complement system. Deeply understanding how the main governing parameters change with respective to the different combinations of pathway activation will greatly aid in development of drugs for strategic therapeutic targets. Due to the low computational cost relative to the existing models and accuracy of our predictions, we believe that our reduced order complement network is the first step towards building a computation toolbox for treatment analysis or clinical screening in the medical field.

@@

\subsection*{Formulation of a reduced order complement model}
We developed a reduced order human complement network consisting of the most crucial steps of the human complement system (Fig. \ref{fig-schematic}). The core of our model was based upon the experimental measurements of Morad and coworker's earlier work \cite{morad2015time}, we only consider the activation of complement system through the alternate and the lectin pathways. In doing so we aim to capture a complex biological phenomenon using a few simple ordinary differential equations.  A trigger event initiates the lectin pathway in the presence of zymosan, which activates the cleavage of C2 and C4 into C2a and C2b, and C4a and C4b respectively. Classical Pathway (CP) C3 convertase (C4aC2b) is a combination of C4a and C2b, which catalyzes the cleavage of C3 into C3a and C3b. Similarly, the activation of the alternative pathways happens through the spontaneous hydrolysis of C3 which facilitates the cleavage of C3. C3b then could combine with with C3 to form alternate pathway (AP) C3 convertase. Both C3 convertases catalyze the cleavage of C3 into C3a and C3b, and C3b can then combine with either CP or AP C3 convertase to form C5 convertase, CP or AP respectively that is responsible for the cleavage of C5 to C5a and C5b. Lectin pathway activation was approximated using a combination of saturation kinetics and Hill-like function control functions. These control coefficients then modified the rates of model processes at each time step. Hill-like transfer functions $0 \le \mathbf{f (Z)}  \le 1$ quantified the contribution of components upon a target process, in this study, $\mathbf{Z}$ represents the abundance of the initiator. Taken together, while the reduced order human complement model encodes significant biological complexity, it is highly compact (consisting of only 18 differential equations). Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.

\subsection*{An ensemble of complement models was estimated using dynamically dimensioned search.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. We estimated
kinetic and control parameters in a hierarchical fashion using  two $in~vitro$ time-series human complement data sets with and without zymosan present. The residual between simulation and experimental measurements were minimized  using dynamically dimensioned search (DDS). An initial parameter set was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized the residual. Knowing that the kinetic and control parameters of the lectin pathway does not affect the dynamics of the alternate pathway, we used a hierarchal approach that estimated the parameters for the alternative pathway and lectin pathway separately. For the alternative pathway, we utilized the time-course experimental measurements of Morad and coworkers \cite{morad2015time} of $C3a$ and $C5a$ in the absence of zymosan and only allowed the alternative parameters to vary (Fig. \ref{fig-fit} A and B). The estimated alternate parameters was then fixed for the determination of lectin pathway parameters. The training for the lectin parameters, we used the experimental measurements of $C3a$ and $C5a$ in the presence of $1$ g of zymosan published by Morad et al \cite{morad2015time} (Fig. \ref{fig-fit} C and D).
The reduced human complement model captured the behavior of the alternative and lectin pathways through the time-course abundance of C3a and C5a (Fig. \ref{fig-fit}).
However we were not able to capture the curvature of the C5a alternative (Fig. \ref{fig-fit}).
The decreasing slope of the experimental measurements may be an indication of the decreasing cofactors that are required for the spontaneous hydrolysis in the alternative pathway, which we neglected.  Taken together, the model identification results suggested that our reduced order approach could reproduce a panel of lectin pathway initiation data sets in the neighborhood of physiological factor and inhibitor concentrations. However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.

We tested the predictive power of the reduced order human complement model with validation data sets not used during model training. Six validation data sets were used, three for $C3a$ and $C5a$ respectively at different zymosan concentrations. All kinetic and control parameters were
fixed for the validation simulations.
The reduced order model predicted the $C3a$ and $C5a$ time-course profiles at a qualitative level (Fig. \ref{fig-prediction}).

@@

In this study, we present an effective biochemical network modeling approach in building a reduced order model to study a complex biochemical network, the human complement system. The key innovation of our approach is the use of simple equations to capture the behavior of a complex biochemical network.  We developed a hybrid modeling approach which combines ODEs and logical rules to model biochemical processes for which a complete mechanistic understanding is missing. Using this modeling framework we  incorporated computational simulation and experimental validation of the lectin and alternative pathway to examine C3 and C5 convertase formation and activity. The reduced order model consisted of only 18 differential equations with 28 kinetic and control parameters. Thus, the model was several orders of magnitude smaller and includes more pathways than comparable purely ODE models in the literature. We estimated the model parameters from in vitro human complement time-course experiments of C3a and C5a from Morad and coworkers \cite{morad2015time}, in the presence and absence of zymosan, and without the classical pathway. We then compared the model predictions with C3a and C5a data sets, for alternative and lectin pathway dynamics, that were not used for model training. Once validated, we performed sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. The reduced order hybrid approach produced a surprisingly predictive human complement model, much similar to the study on human coagulation using the same modeling framework \cite{sagar2015dynamic}. Taken together, the combined analysis of alternate and lectin pathways along with the incorporation of the downstream reactions involving C5 convertase elucidated new insight into roles of governing parameters that mediate the complement system. Deeply understanding how the main governing parameters change with respective to the different combinations of pathway activation will greatly aid in development of drugs for strategic therapeutic targets. Due to the low computational cost relative to the existing models and accuracy of our predictions, we believe that our reduced order complement network is the first step towards building a computation toolbox for treatment analysis or clinical screening in the medical field.

The performance of the reduced order complement model was impressive given its limited size. However, there are several critical questions that should be explored following this study. A logical progression for this work would include expanding the network to include the classical pathway and the formation of the membrane attack complex (MAC). It is unclear whether the addition of the classical pathway will decrease the prediction of our existing model due to the cross-talk between the classical and lectin activation shown by Liu et al \cite{liu2011computational}. One potential approach in addressing such difficulties would be the incorporation of additional species such as C reactive proteins (CRP) and L-ficolin (LF) that involved in complement initiation of classical and lectin pathways. The influence of CRP, LF and the cross-talk can be captured through additional control functions that act upon the initiation pathways in a logical integration rule developed by Wayman and coworkers \cite{wayman2015dynamic}. Another issue with our reduced order model involve the omitted species that are implicitly lumped together with our effective kinetics and control parameters. Due to the reduction of parameters, the model cannot determine the dynamics or explicit impact of the omitted species on the system. However, we have created a hierarchy approach for parameter estimation that can be used to uncouple the kinetic parameter and contribution of any additional complement proteins and regulators. Using this simple and versatile modeling approach that we created, we took the first step in the development of a computation toolkit that can be readily used in a clinical setting. Our reduced order complement model is computationally inexpensive, and versatile so it could easily be incorporated into pre-existing or new pharmacokinetic models. Furthermore this approach model has the potential to create individualized treatment plans for patients with complement deficiency.

@@

\section*{Supplemental materials.}
\subsection*{Model equations.}
The reduced-order complement model consisted of 18 ordinary differential equations, 12 rate equations, and two control equations:
\begin{eqnarray}
	\frac{dx_{1}}{dt} & =& - r_{1}f_{1} \\ 								%C4
	\frac{dx_{2}}{dt} &=& - r_{2}f_{2} \\  								%C2
	\frac{dx_{3}}{dt} &=&  r_{1}f_{1} \\ 								%C4a
	\frac{dx_{4}}{dt} &=& r_{1}f_{1} - r_{6} \\ 					%C4b
	\frac{dx_{5}}{dt} & = & r_{2}f_{2} - r_{6} \\ 					%C2a
	\frac{dx_{6}}{dt} &=& r_{2}f_{2} \\ 								%C2b
	\frac{dx_{7}}{dt} &=& r_{3} - r_{4} - r_{5}\\ 				%C3
	\frac{dx_{8}}{dt} &=& r_{3}  + r_{4} + r_{5}  - k_{deg,c3a}*C3a \\		%C3a
	\frac{dx_{9}}{dt} &=& r_{3}  + r_{4} + r_{5}  -  r_{7} \\		%C3b
	\frac{dx_{10}}{dt} &=& r_{6} - r_{10} -  r_{8} \\			%CP C3C
	\frac{dx_{11}}{dt} &=& r_{7} - r_{11} -  r_{9} \\			%AP C3C
	\frac{dx_{12}}{dt} &=& r_{10} - r_{14} \\							%CP C5C
	\frac{dx_{13}}{dt} &=& r_{10}  \\											%AP C5C
	\frac{dx_{14}}{dt} &=& - r_{12} - r_{13} \\								%C5
	\frac{dx_{15}}{dt} &=&  r_{12} + r_{13} - k_{deg,c5a} \\					%C5a
	\frac{dx_{16}}{dt} &=& r_{12} + r_{13} \\								%C5b
	\frac{dx_{17}}{dt} &=& - r_{8} - r_{14} \\  				         %C4BP
	\frac{dx_{18}}{dt} &=& - r_{9} \\									%Factor H
\end{eqnarray}where the rate equations are given by:
\begin{eqnarray}
	r_{1} &=& \frac{k_{i1}(C4)}{(K_{1s} + C4)} \\
	r_{2} &=& \frac{k_{2}(C2)}{(K_{2s} + C2)} \\
	f_{1} &=& \frac{Zymo^{\eta_{1}}}{(Zymo^{\eta_{1}} + \alpha_{1}^{\eta_{1}})} \\
	f_{2} &=& \frac{Zymo^{\eta_{2}}}{(Zymo^{\eta_{2}} + \alpha_{2}^{\eta_{2}})} \\
	r_{3} &=& k_{3}(C3) \\
	r_{4} &=& \frac{k_{4}(C3C_{L})(C3^{\eta_{3})}}{(K_{4s}^{\eta_{3}} + C3^{\eta_{3}})}  \\
	r_{5} &=& \frac{k_{5}(C3C_{A})(C3)}{(K_{5s} + C3)} \\
	r_{6} &=& k_{6}(C4b)(C2a) \\
	r_{7} &=& k_{7}(C4b)(C2a) \\
	r_{8} &=& k_{8}(C3C_{L})(C4b)(C4BP) \\
	r_{9} &=& k_{9}(C3C_{A})(FactorH) \\
	r_{10} &=& k_{10}(C3C_{L})(C3b) \\
	r_{11} &=& k_{11}(C3C_{A})(C3b) \\
	r_{12} &=& \frac{k_{12}(C5C_{L})(C5^{\eta_{4}})}{(K_{12s}^{\eta_{4}} + C5^{\eta_{4}})} \\
	r_{13} &=& \frac{k_{13}(C5C_{A})(C5)}{(K_{13s} + C5)} \\
	r_{14} &=& k_{14}(C5C_{L})(C4BP)
\end{eqnarray}
